# Mixed models

These exercises extend what was previously introduced in the association tests from regression models to mixed models that account for genetic relatedness among samples.

## Null model

Recall that the first step in our association testing procedure is to fit the null model. In addition to the `AnnotatedDataFrame` with phenotype data that we used previously, we will also use the ancestry PCs and pairwise kinship estimates we created in the previous session. We will use the first 4 PCs to adjust for ancestry.

```{r null_model_mm, message = FALSE}
# sample annotation
repo_path <- "https://github.com/UW-GAC/SISG_2021/raw/master"
if (!dir.exists("data")) dir.create("data")
sampfile <- "data/sample_phenotype_annotation.RData"
if (!file.exists(sampfile)) download.file(file.path(repo_path, sampfile), sampfile)
annot <- get(load(sampfile))
library(Biobase)
head(pData(annot))

# load the ancestry PCs
pcfile <- "data/pcs.RData"
if (!file.exists(pcfile)) download.file(file.path(repo_path, pcfile), pcfile)
pcs <- get(load(pcfile))
pcs <- pcs[,c("sample.id", "PC1", "PC2", "PC3", "PC4")]
head(pcs)

# merge PCs with the sample annotation
library(dplyr)
dat <- left_join(pData(annot), pcs, by="sample.id")
# update the AnnotatedDataFrame
pData(annot) <- dat
save(annot, file="data/sample_phenotype_pcs.RData")
```

We can create a kinship matrix from the output of `pcrelate`. We multiply the kinship values by 2 to get values on the same scale as the standard GRM. This matrix is represented in R as a symmetric matrix object from the Matrix package.

```{r load_kinship}
kinfile <- "data/pcrelate_kinship.RData"
if (!file.exists(kinfile)) download.file(file.path(repo_path, kinfile), kinfile)
pcrel <- get(load(kinfile))
library(GENESIS)
kinship <- pcrelateToMatrix(pcrel, scaleKin=2, verbose=FALSE)
dim(kinship)
kinship[1:5,1:5]
```

When running a mixed model analysis, we still fit the null model using the `fitNullModel` function in GENESIS. Now, we include the kinship matrix in the model with the `cov.mat` argument, which is used to specify the random effect(s) in the model with covariance structure(s) proportional to the supplied matrix(s). The inclusion of these random effects is what makes this a mixed model, rather than a simple regression model. We also add the ancestry PCs to the list of covariates and allow for heterogeneous residual variance by `study` with the `group.var` argument, as before.

```{r null_model_fit_mm}
nullmod <- fitNullModel(annot, 
                        outcome="height", 
                        covars=c("sex", "age", "study", paste0("PC", 1:4)), 
                        cov.mat=kinship, 
                        group.var="study", 
                        verbose=FALSE)
save(nullmod, file="data/null_mixed_model.RData")
```

We can investigate the output from `fitNullModel`.
```{r}
# description of the model we fit
nullmod$model

# fixed effect regression estimates
nullmod$fixef

# variance component estimates by group.var
nullmod$varComp

# model fit: fitted values, residuals
head(nullmod$fit)

library(ggplot2)
ggplot(nullmod$fit, aes(x = fitted.values, y = resid.marginal)) + 
    geom_point(alpha = 0.5) + 
    geom_hline(yintercept = 0) + 
    geom_smooth(method = 'lm')
```

## Single-variant tests

Now we can run a single-variant test, accounting for genetic ancestry and genetic relatedness among the subjects. We use the same `assocTestSingle` function as before; the only difference is that we pass in our new null model. 

```{r assoc_single_mm, message = FALSE}
library(SeqVarTools)
gdsfile <- "data/1KG_phase3_subset_chr1.gds"
if (!file.exists(gdsfile)) download.file(file.path(repo_path, gdsfile), gdsfile)
gdsfmt::showfile.gds(closeall=TRUE) # make sure file is not already open
gds <- seqOpen(gdsfile)

# make the seqVarData object
seqData <- SeqVarData(gds, sampleData=annot)

# make the iterator object
iterator <- SeqVarBlockIterator(seqData, verbose=FALSE)

# run the single-variant association test
assoc <- assocTestSingle(iterator, nullmod)
dim(assoc)
head(assoc)
```

We make the usual QQ plot, filtering to variants with minor allele count (MAC) >= 5.

```{r assoc_single_qq_mm}
library(ggplot2)
qqPlot <- function(pval) {
    pval <- pval[!is.na(pval)]
    n <- length(pval)
    x <- 1:n
    dat <- data.frame(obs=sort(pval),
                      exp=x/n,
                      upper=qbeta(0.025, x, rev(x)),
                      lower=qbeta(0.975, x, rev(x)))
    
    ggplot(dat, aes(-log10(exp), -log10(obs))) +
        geom_line(aes(-log10(exp), -log10(upper)), color="gray") +
        geom_line(aes(-log10(exp), -log10(lower)), color="gray") +
        geom_point() +
        geom_abline(intercept=0, slope=1, color="red") +
        xlab(expression(paste(-log[10], "(expected P)"))) +
        ylab(expression(paste(-log[10], "(observed P)"))) +
        theme_bw()
}    

qqPlot(assoc$Score.pval[assoc$MAC >= 5])
```

Notice that we observe much less inflation than before, when we did not adjust for ancestry and relatedness.

## Exercise

1. Perform a single-variant association test for `status`. Adjust for sex, age, study, ancestry, and kinship in the model. Don't forget to consider the `family` parameter. Make a QQ plot of the p-values for all variants with MAC >= 5.


## Solution

1. Perform a single-variant association test for `status`. Adjust for sex, age, study, ancestry, and kinship in the model. Don't forget to consider the `family` parameter. Make a QQ plot of the p-values for all variants with MAC >= 5.

The first step is to fit the null model. We include the first 4 PCs as covariates in our model to adjust for ancestry, and we include a random effect proportional to the kinship matrix to adjust for genetic relatedness. Recall that with a binary outcome, we set `family = binomial` -- because we have a random effect, this will fit an approximate logistic mixed model using the [GMMAT method](https://www.cell.com/ajhg/fulltext/S0002-9297(16)00063-X).

```{r exercise_mm_nullmod}
nullmod.status <- fitNullModel(annot, 
                               outcome="status", 
                               covars=c("sex", "age", "study", paste0("PC", 1:4)), 
                               cov.mat=kinship, 
                               family = binomial,
                               verbose=FALSE)

# description of the model we fit
nullmod.status$model

# fixed effect regression estimates
nullmod.status$fixef

# variance component estimates by group.var
nullmod.status$varComp
```

Now that we have the null model, we perform the single-variant association tests and make the QQ plot the same way as before. 

```{r exercise_mm_assoc}
resetIterator(iterator, verbose=FALSE)
# run the single-variant association test
assoc.status <- assocTestSingle(iterator, nullmod.status, test="Score")
dim(assoc.status)
head(assoc.status)

qqPlot(assoc.status$Score.pval[assoc.status$MAC >= 5])
```

```{r assoc_mm_close}
seqClose(seqData)
```
